Generalising the impact of forest cover on streamflow from experimental data: it is not that simple.

### 2022-09-28 paper/Forest_and_Water/Reply_first_review_JoH.Rmd

---
title: "Replies to editor and associate editor Journal of Hydrology"
author:
  - name: R. Willem Vervoort
    correspondingauthor: true
  - name: Eliana Nervi
  - name: Jimena Alonso
date: "`r Sys.Date()`"
linenumbers: true
#numbersections: true
csl: elsevier-harvard.csl
output:   
  bookdown::pdf_book:
    base_format: rticles::elsevier_article
bibliography: forestandwater.bib
editor_options: 
  chunk_output_type: inline
header-includes:
  - \usepackage{setspace}\doublespacing
  - \usepackage{color}
abstract: |
  This document records the replies to the reviewer for the first submission of *Factors determining how catchments respond to forest cover change. Re-analysing global data sets* to Journal of Hydrology, which was rejected before review.
---

@@

```{r setup, include=FALSE, echo =F}
knitr::opts_chunk$set(echo = TRUE)
require(tidyverse)
require(lubridate)
suppressWarnings(require(knitr))
suppressWarnings(require(bookdown))
suppressWarnings(require(pander))
```

# Introduction

The reviewer comments are ordered with the Editor in Chief comments listed first and our responses next. This is followed by the comments from the associate editor and our responsed. The reviewer comments are in \textcolor{blue}{blue} and our responses in normal text.

@@

# Editor comments

\textcolor{blue}{The first is to streamline the statistics.  As suggested by the AE, a formal model selection process, followed by using only the selected model(s) to evaluate change, would be a suitable approach.}

We politely disagree with the suggestion of a formal model selection process, as we outline in our reply to the AE. If the statistical modelling was aimed at developing the best predictive model, then this would be the right approach. However, in this case the statistical modelling is aimed at hypothesis testing and explanation of variance in the data set.

@@

\textcolor{blue}{The second - hopefully supported by a streamlining of the statistics - is to simplify and improve the coherence of the argument.  Framing - as the AE states - a "systematic" exploration of the importance of parameters, will improve the readability and interpretability of the work.  At present, I think because the statistical analysis itself is convoluted - so too the thread of the argument and clarity of the messages are hard to follow.}

Please see our comments in reply to the AE below.

\textcolor{blue}{The final - and perhaps in truth the most problematic issue that may stand in the way of the MS being published - relates to novelty.  The AE highlights several recent works with similar questions, approaches and findings at global scales.  I could add to that list with more regional studies (e.g. Levy et al 2018).  So carefully identifying the knowledge gap being addressed, with respect to these recent studies, and making the case for the present study being "needed" will also be essential.}

Thank you for the suggestion of the @levy2018 paper, which is an excellent example of a careful statistical analysis taking into account possible variations in climate and dynamic land use change. This is exactly the kind of statistical analysis that we believe needs to be undertaken to better understand how forest cover impacts  The paper also provides the rainfall and runoff data that was used in the study, but regrettably does not provide the land use data. Rather than deriving this ourselves following the methods described in the paper, we decided that it would be better to use @levy2018 as an example of how the analysis can be done well.

You are quite right that we need to be clearer about the objectives of the paper and how this is different from the existing work. We agree that there have been many attempts to derive general conclusions in relation to the impact on streamflow of changes to forest cover, as we also discuss in our introduction. However, as we discuss in our response to the Associate Editor, there are considerable issues with the generalization of such studies (as attempted in @zhou2015; @jackson2005; @filoso2017 and @zhang2017). This is now extensively discussed in the discussion of the paper.

@@

We agree that outliers could affect the observed residual distribution as this would most likley be obvious in the tails of the distribution. As @venables2013modern outline on p119:
"Outliers are sample values that cause surprise in relation to the majority of the sample. This is not a pejorative term; outliers may be correct, but they should always be checked for transcription errors." As a result, we believe that excluding values that are outliers is probably not a good idea.

However, another careful review of the data identified many further errors in the data, which were all originally in @zhang2017. A particular problem was that many catchments had the wrong sign for the change in forest cover. There are many catchments with reported positive change in cover and a large increase in flow. These were all checked and corrected if needed and a full list of all these changes is below and is now included in the paper as Supplementary Data Part 1:

- 76, Beaver Creek, the flow was corrected from 600% to 157% after review of the original publication [@baker1984].
- 124, D3, @amatya2008effects: The originally recorded 250\% change by @zhang2017 is clearly wrong. The paper says on page 7: Both of these outflow ratios (0.64 and 0.50) were higher than the calculated expected values of 0.55 for 2003 and 0.44 for 2005, respectively. So value should be $0.64/0.55*100 - 100$ or $0.5/044*100 - 100$: 16\% or 13\%. corrected to 16\%
- 3, Baker Creek, @zhangwei2012. The original recorde 201.1\% change by @zhang2017 is also wrong. Original paper says on page 2031: Annual mean flow has been increased by 47.6\%. corrected.

@@

## Comment AE 3

\textcolor{blue}{- The manuscript presents a number of alternative statistical models, differing by candidate explanatory variables. Each model is designed considering the key shortcomings of the previous one. The end result of such an approach is a complex and somewhat non systematic exploration of predictors and their explanatory power, where it is easy to get lost. I suggest restructuring the manuscript around a well-designed and robustly formalized model selection. One way to proceed could be to start with the most complex model suggested by the extant understanding of the processes at play, and then proceed with a model simplification, according to some consistent criteria (AIC, dropping non significant terms, or similar; high r2 is not a good criterion because it does not consider the number of parameters). A full blown model selection would also allow to retain or discard the interaction terms, which could be important (as also recognized by the authors; Section 4.5) and should not be discarded a priori. Doing a proper model selection and presenting the results only for the best model (according to a clearly specified criterion) would be less subjective and allow to drastically reduce the number of figures and tables, allowing the reader to focus their attention on the key message.}

<!-- answer -->
The Associate editor raises several points that need further discussion.

- a formalised model selection;
- the use of AIC rather than *r*^2^ for model selection; and
- interaction terms.

@@

### A formalised model selection

We agree that a classical statistical approach would involve a formal model selection. We acknowledge that in the manuscript we ended up mixing two approaches, where we should have stayed with a single approach. As outlined earlier, the current focus of the statistical modelling is on understanding the different covariates that explain the variation in the data, and to identify latent variables that cause the apparent relationships in the data. This means that there is no attempt to do a formal model selection, as we are not seeking the best predictive model.

However, we have written this badly in the original manuscript and have now reworded large sections of the methodology and the results to be more clear about this approach (line 220 - 224 on page 11)

### AIC rather than *r*^2^

We actually used the adjusted *r*^2^ in the paper, which does take into account the number of degrees of freedom in the model and therefore can be used to discriminate between models, in exactly the same way as the AIC does. However, as the GAM models also provide an AIC, we have for consistency changed our model performance measure to the AIC.

### 2022-09-28 paper/Forest_and_Water/Forest_and_Water.Rmd
revision of draft for internal review

---
title: 'Generalising the impact of forest cover on streamflow from experimental data: it is not that simple.'
author:
  - name: R. Willem Vervoort
    email: willem.vervoort@sydney.edu.au
    affiliation: a,b
    correspondingauthor: true
  - name: Eliana Nervi
    email: eliananervif@gmail.com
    affiliation: c
  - name: Jimena Alonso
    email: jalonso@fing.edu.uy
    affiliation: d
address:
  - code: a
    address: The University of Sydney, Sydney, NSW 2006, Australia
  - code: b
    address: ARC ITTC Data Analytics for Resources and Environments, Sydney Institute of  Agriculture, School of Life and Environmental Sciences.
  - code: c
    address: Project Manager, FPTA 358, Instituto Nacional de Investigacion Agropecuaria,  INIA-Uruguay, Ruta 48 km 10, Rincon del Colorado, 90100 Canelones, Uruguay
  - code: d
    address: Institute of Fluid Mechanics and Environmental Engineering, School of Engineering, Universidad de la Republica, 11200 Montevideo, Uruguay
abstract: |
  Three recent papers review and analyse large global datasets related to impacts of forest cover on streamflow. Using three different approaches, they all find a strong relationship between forestation, de-forestation and streamflow. The data for the recent three papers were reviewed, combined and re-analysed using generalised additive modelling. The overall results of the modelling indicates similar results to previous papers. The results are however problematic, the data set is unbalanced, and there are underlying correlations that suggest further factors that influence te results. The area of the catchment is strongly related to the assessment technique and the variability in the response data.
  A careful review of the data and the methodology in the previous papers highlights that there are  four interlinked reasons that are problematic:  1) The existence of latent variables in the data that create the appearance of a relationship that really does not exist; 2) The difficulty in fully interpreting the specifics of different studies; 3) The difficulty of integrating data from seemingly similar studies, but with quite different objectives; and 4) The chance of transcription errors influencing the data. Overall this indicates that while valuable data can be extracted from past studies, the above problems need to be considered before results are extrapolated to greater and global scales.  
keywords: 
  - data analysis
  - forest cover change
  - global scales
  - literature review
journal: "Journal of Hydrology"
date: "`r Sys.Date()`"
# classoption: preprint, 3p, authoryear
bibliography: forestandwater.bib
linenumbers: true
#numbersections: true
csl: elsevier-harvard.csl
output:   
  bookdown::pdf_book:
    base_format: rticles::elsevier_article
header-includes:
    - \usepackage{setspace}\doublespacing
---

@@

There is a urgent need to identify the impacts of human intervention on streamflow at a global scale and to separate this from climate effects [@wang2020; @hoekvandijke2022]. More specifically, the impacts of global deforestation and reforestation are important through their perceived influence on streamflow and blue and green water availability [@hoekvandijke2022; @schyns2019].  The past work reviewing these impacts [@andreassian2004; @jackson2005; @zhang2017; @brown2005; @brown2013; @filoso2017] highlights a general consensus that if forest areas increase, streamflow decreases and vice-versa. The most dramatic result is Figure 5 in @zhang2011 indicating (for Australian catchments) a 100% decrease in streamflow for catchments with 100% forest cover. However, on the other end of the spectrum, for three French catchments [@cosandey2005], there was no change in streamflow characteristics in two of the catchments after deforestation. For reforestation, a modelling study across the 1 millon km^2^ Murray Darling Basin also found no major effect, especially in larger catchments [@vandijk2007], but a recent study [@hoekvandijke2022] found an 8\% changes in streamflow as a result of reforestation. Similarly a modelling study by @beck2013 found no significant change in streamflows in 12 catchment in Puerto Rico as a result of deforestation. In contrast, in a recent study in Brazil across 324 catchments, @levy2018 found a significant increase in streamflow, particular in the dry season, as a result of deforestation. This suggests that there can be significant variation across the different studies,methodologies and geographical regions.

@@

As mentioned, several review papers have summarized the plethora of forestation and deforestation studies across the globe, in relation to paired watershed studies [@brown2005; @hewlett1984], related to reforestation in particular [@filoso2017], and more generally [@jackson2005; @zhang2017].  These studies aim to generalize the individual experimental and research findings and to identify if there are global trends or relationships. Others have used the understanding from these studies to extrapolate to global scales [@hoekvandijke2022].

There has been a recent push in the hydrological community [@evaristo2020metaanalysis] to use 'meta-analysis' to summarise past studies. The claim is that because a meta-analysis uses clearly defined search terms and statistical methods to analyse the results, that this will lead to more reliable summaries of past research.

From the most recent reviews on forest impact on streamflow [@zhang2017; @filoso2017], the study by @filoso2017 is a clear meta-analysis, while the other is not. However, both authors have developed an impressive global database of catchment studies with changes in streamflow due to changes in forest coverand use statistical approaches to analyse the resulting data. The @zhang2017 dataset, which covers over 312 studies, is described in terms of the change in streamflow as a result of the change in forest cover, where studies related to both forestation (increase in forest cover) and deforestation (decrease in forest cover) were included. In contrast, the paper by @filoso2017 focused primarily on reforestation, and covered an equally impressive database of 167 studies using a systematic review. In this case the collected data is mostly coded as count data and only a subset of 37 studies was analysed for actual water yield change. There is some overlap between the two data sets, but there are also some studies unique to both sets. The more regionally concentrated and detailed study by @levy2018 is a further independent dataset with no overlap with the other studies. However, for this study only the flow and rainfall data is available for the catchments, and the change in landcover was derived from satellite data and was not made available.

The conclusions of the first mentioned major review paper [@zhang2017] indicates that there is a distinct difference in the change in flow as a result of forestation or deforestation between small watersheds (catchments), defined as < 1000 km^2^ and large watersheds (catchments) > 1000 km^2^. While for small catchments there was no real change in runoff with changes in cover, for large catchments there was a clear trend showing a decrease in runoff with and increase in forest cover. Their main conclusion was that the response in annual runoff to forest cover was scale dependent and appeared to be more sensitive to forest cover change in water limited catchments relative to energy limited catchments [@zhang2017].

@@

A final earlier summary paper that includes much of the same data as @zhang2017 and @filoso2017 is @zhou2015, which has one author in common with @zhang2017. However, this paper aims to explain the variation in the data using the elasticity approach in the Fuh model. In particular, it aims to link the variation in the observed data to variations in the exponent *m* in the Fuh model. A key observation is that in drier environments, the effects of removing forest cover are much greater than in wetter environments, which is also suggested by Figure 4 in @zhang2017. The Fuh model and variations of the Budyko equilibrium modelling approach was also used by @hoekvandijke2022 to interpret the global impact of reforestation.

There are some clear limitations in these studies, and some of this applies to meta-analyses in general. The main method in the work by @zhang2017 is a single covariate linear regression. In contrast, the systematic review from @filoso2017 emphasises the classification and distributions of the study. @zhang2017 points out that a main assumption in their work is that the catchment size threshold at 1000 km^2^ is a distinct separation between "small" and "large" catchments. However, a subset of 37 data points in @filoso2017 (their Figure 9) does not appear to support this, suggesting a continuum. And while the work @filoso2017 provides important insights in study types, analysis types, forest types and broad classification, there is limited quantification of actual impact.

In contrast to the single covariate linear regression in the earlier studies [@zhang2017; @filoso2017] and the top-down Budyko modelling [@zhou2015; @hoekvandijke2022], the regional Brazilian Cerrado study [@levy2018] provides a carefully designed statistical approach using mixed effects modelling and Differences-in-Differences modelling focusing specifically on the effect of deforestation. The analysis specifically accounted for differences between catchments and differences due to variations in climate. Their conclusion highlighted that in particular dry season streamflow was affected by deforestation.

Given all these previous reviews and the seemingly clear conclusions about the impact of forest cover change on streamflow, the question is why another paper?
There is a real attraction in the meta-analysis idea of statistical analysis of past studies to be able to extrapolate findings to larger scales and to identify factors across global scales [@evaristo2020metaanalysis]. 
However, there is also a real danger in this process, which this paper aims to highlight. There are four potential errors (or limitations) in such global meta-analyses:

@@

The aim of this paper is to first reanalyse the global dataset using some more sophisticated statistical modelling and uses this to highlight examples of each of these limitation. This will show how they have influenced the outcomes of the past work, and provide suggestions of how we can overcome these limitations. In addition, by applying more complex statistical models we will highlight the conclusions that can still be drawn from this work in relation to the impact of forest cover on streamflow. Finally, we will highlight future research needs in this area.

@@

As indicated, the starting point of this paper is the data base of studies which were included in @zhang2017 as supplementary material. The columns in this data set (are the catchment number, the	catchment name, the	Area in km^2^, the annual average precipitation	(Pa) in mm, the	forest type, hydrological regime, and climate type, the change in forest cover	in \% ($\Delta F\%$) and the change in streamflow in \% $(\Delta Qf\%)$, based on equation 1 in @zhang2017), the precipitation data type, the assessment technique, and the	source of the info, which is a citation. 
Several of these columns contain abbreviations to describe the different variables, which are summarised in Table \@ref(tab:table1). These abbreviations will later be used in the models.

@@

The paper by @zhang2017 use the dryness index, which is the annual rainfall (Pa) divided by the potential or reference evapotranspiration (ET~0~ or E~0~) in their analysis, and use the dryness index to identify the climate type. However, the potential or reference ET was not originally included as part of the published data set. We combined the tables for small catchments (< 1000 km^2^) and large catchments (>= 1000 km^2^) from @zhang2017 in our analysis.

@@

As highlighted, @zhang2017 did not provide values for evapotranspiration in the data base. Using the location information  reference evapotranspiration (E~0~) was extracted from the Global Aridity Index and Potential Evapo-Transpiration (ET~0~) Climate Databasev2 [@trabucco2018], if a value of E~0~ was not available from the original papers. For large catchments, this value (and the associated coordinates), similar to annual average rainfall, is only an approximation of the climate at the location.

Similar to @zhang2017, the "dryness index" was calculated from the reference evapotranspiration and the annual average rainfall (Pa) as:

\begin{equation}
D = \frac{E_{0}}{Pa} (\#eq:eq1)
\end{equation}

@@

The aim of the statistical analysis is to highlight the most important variables in the data set that explain the change flow as a consequence of changes in forest cover. This first aim is similar to @zhang2017, but the main difference is that we start off with all variables in the data set in the model. Subsequently the analysis will concentrate on how the individual variables in the dataset relate to each other and  how latent variables in the data set can be masked and result in relationships that might not really exist. Finally, the analysis will highlight how the results are conditional on the dataset.

In the statistical analysis we are not necessarily seeking the best "predicitive" model, and as such do not perform a traditional variable selection process. Rather, we focus on analyzing the predictor variables in the full model to identify how all the variables explain the variance in the dependent variable.

@@

The changes in forest cover contain both positive (forestation) and negative values (deforestation). In @zhang2017, these changes were jointly analysed, assuming the effect on the change in flow was linear and the effect of removing forest cover was the same as an equivalent addition of forest cover.

However, the impact of an increase in forest cover can be different from the same fractional decrease in forest cover. The question becomes how best to analyse this. One approach would be to allow a different slope and a different intercept for the decreases relative to the increases. 
This can be tested by converting all the change in forest cover data to positive values, and an additional binary column ($sign_{forest cover}$) can be included indicating whether it was a forest cover increase or decrease. In the model, the parameter for $sign_{forest cover}$ will indicate the difference in the changes in flow for increases in forest cover compared to decreases in forest cover. The disadvantage of this approach is that the relationship with forest cover becomes discontinuous at the origin (0 change in forest cover).

A second approach is to test the change in forest cover as a non-linear relationship in the GAM model. Because a shrinkage penalty is used, this will also test the non-linear assumption and allows the variable for forest cover to be continuous. The disadvantage of this approach is that the relationship between forest cover and change in flow is less easy to interpret, as the non-linear fit in the GAM has no direct parametric form. 
Both these approaches are tested in the results.

The over arching test focuses on identifying the change streamflow as a result of a change in forest cover and potentially affected by different other factors (as indicated by the previous research: @zhang2017; @filoso2017; @zhou2015): climate, size of catchment and length of study. In addition to these earlier identified factors, this study also tested for the factors listed in Table \@ref(tab:table1)

@@

The overall dataset contains `r nrow(All_data)` observations of changes in flow, which includes the newly identified data sets and after removing identified duplicate data and lines with missing data. In contrast, the original dataset from @zhang2017 contained `r nrow(Zhang_all)` catchments and the @filoso2017 study used 37 catchments (Table S2 in @filoso2017). The current number of catchments is the result of the removal of duplicates and our modifications and additions. The overall distribution of changes in flow is highly skewed as is the distribution of changes in forest cover and *Area km^2^*. The values of changes in flow greater than 100\% and smaller than -100\% clearly create long tails on the change in flow distribution. Note also the large number of studies with 100\% forest cover reduction. Clearly visible is also that smaller catchments dominate the database with 42\% of the data from catchments < 1 km^2^ and 65\% of the data for catchments < 10 km^2^ (Figure 1). This high skew in some of the data can create difficulties in the statistical modelling and further transformation of the data might be required.

```{r, echo = F, warning = F, message=F, eval =F}
png("DataExploration.png", width = 2500, height = 1800, res = 300)
All_data %>%
  mutate(log10Area = log10(Area_km2)) %>%
  pivot_longer(c(DeltaF_perc, DeltaQf_perc, log10Area, Pa_mm, E0), 
               names_to = "variable", values_to = "value") %>%
  ggplot(aes(value)) + geom_histogram(fill = "steelblue", bins=50) + theme_bw() +
  facet_wrap(~variable, scales = "free", labeller = labeller(variable = 
    c("log10Area" =  "log10(Area km2)", #latex2exp::TeX("$\\log_{10}({Area~km^2})$"),
      "DeltaF_perc" = "Delta Forest (%)",
      "DeltaQf_perc" = "Delta Qf (%)",
      "E0" = "Annual Potential E (mm)",
      "Pa_mm" = "Annual Rainfall (mm)"),default = label_parsed)) +   
  theme(axis.title = element_text(size = rel(1.5)),
        axis.text = element_text(size = rel(1.2)),
        strip.text = element_text(size = rel(1.5)),
        axis.text.x = element_text(angle = 45, hjust = 1))
dev.off()
```

@@

```{r correlation, eval=F, echo = F}
All_data2 <- All_data2 %>%
  mutate(length = To - From,
         mid_year = From + (To - From)/2)

>>>>

---
title: 'Generalising the impact of forest cover on streamflow from experimental data: it is not that simple.'
author:
  - name: R. Willem Vervoort
    email: willem.vervoort@sydney.edu.au
    affiliation: a,b
    correspondingauthor: true
  - name: Eliana Nervi
    email: eliananervif@gmail.com
    affiliation: c
  - name: Jimena Alonso
    email: jalonso@fing.edu.uy
    affiliation: d
address:
  - code: a
    address: Sydney Institute of  Agriculture and School of Life and Environmental Sciences, The University of Sydney, Sydney, NSW 2006, Australia
  - code: b
    address: ARC ITTC in Data Analytics for Resources and Environments, The University of Sydney, NSW 2006, Sydney, Australia
  - code: c
    address: Project Manager, FPTA 358, Instituto Nacional de Investigacion Agropecuaria,  INIA-Uruguay, Ruta 48 km 10, Rincon del Colorado, 90100 Canelones, Uruguay
  - code: d
    address: Institute of Fluid Mechanics and Environmental Engineering, School of Engineering, Universidad de la Republica, 11200 Montevideo, Uruguay
abstract: |
  Three recent papers review and analyze large global datasets related to impacts of forest cover on streamflow. Using three different approaches, they all find a strong relationship between forestation, de-forestation and streamflow. However, the results are problematic, the underlying data set is unbalanced, and there are correlations in the data that warrant further investigation as this would influence the results. For example, the area of the catchment is strongly related to the assessment technique and the variability in the response data. For this study, the data for the recent three papers were reviewed, combined, and supplemented with new studies. Subsequently, the data were re-analyzed using generalised additive modelling. 
  The results highlight that there are  four interlinked reasons that make the general outcomes from the previous papers problematic:  1) The existence of latent variables in the data that create the appearance of a relationship that really does not exist; 2) The difficulty in fully interpreting the specifics of different studies; 3) The difficulty of integrating data from seemingly similar studies, but with quite different objectives; and 4) The chance of transcription errors influencing the data. Overall this indicates that while valuable data can be extracted from past studies, the above problems need to be considered before results are generalised and extrapolated to continental and global scales.  
keywords: 
  - data analysis
  - forest cover change
  - global scales
  - literature review
journal: "Journal of Hydrology"
date: "`r Sys.Date()`"
# classoption: preprint, 3p, authoryear
bibliography: forestandwater.bib
linenumbers: true
#numbersections: true
csl: elsevier-harvard.csl
output:   
  bookdown::pdf_book:
    base_format: rticles::elsevier_article
header-includes:
    - \usepackage{setspace}\doublespacing
---

@@

There is a urgent need to identify the impacts of human intervention on streamflow at a global scale and to separate this from climate effects [@wang2020; @hoekvandijke2022]. More specifically, the impacts of global deforestation and reforestation are important through their perceived influence on streamflow and blue and green water availability [@hoekvandijke2022; @schyns2019].  The past work reviewing these impacts [@andreassian2004; @jackson2005; @zhang2017; @brown2005; @brown2013; @filoso2017] highlights a general consensus that if forest areas increase, streamflow decreases and vice-versa. The most dramatic example of this is Figure 5 in @zhang2011 indicating (for Australian catchments) a 100% decrease in streamflow for catchments with 100% forest cover. However, on the other end of the spectrum, for three French catchments [@cosandey2005], there was no change in streamflow characteristics in two of the catchments after deforestation. For reforestation, a modelling study across the 1 millon km^2^ Murray Darling Basin also found no major effect, especially in larger catchments [@vandijk2007], but a recent study [@hoekvandijke2022] found an 8\% change in streamflow as a result of reforestation. Similarly a modelling study by @beck2013 found no significant change in streamflow in 12 catchments in Puerto Rico as a result of deforestation. In contrast, in a recent study in Brazil across 324 catchments, @levy2018 found a significant increase in streamflow, particular in the dry season, as a result of deforestation. This suggests that there can be significant variation across the different studies, methodologies and geographical regions.

@@

There has been a recent push in the hydrological community [@evaristo2020metaanalysis] to use 'meta-analysis' to summarise past studies. The suggestion is that, because meta-analyses use clearly defined search terms and statistical methods to analyze the results, this will lead to more reliable summaries of past research. As a result, several review papers have summarized the plethora of forestation and deforestation studies across the globe, in relation to paired watershed studies [@brown2005; @hewlett1984], related to reforestation in particular [@filoso2017], and more generally [@jackson2005; @zhang2017].  These studies aim to generalize the individual experimental and research findings and to identify if there are global trends or relationships. Others have used the understanding from a global analysis to extrapolate to global scales [@hoekvandijke2022].

The recent paper by @filoso2017 is a clear meta-analysis, but most others [@zhang2017; @hoekvandijke2022; @zhou2015] are not. However,  n impressive global database of catchment studies with changes in streamflow due to changes in forest cover has been developed [@zhang2017; @filoso2017] and statistical approaches are used to analyze the resulting data. The @zhang2017 dataset, which covers over 312 studies, is described in terms of the change in streamflow as a result of the change in forest cover, where studies related to both forestation (increase in forest cover) and deforestation (decrease in forest cover) were included. In contrast, the paper by @filoso2017 focused primarily on reforestation, and covered an equally impressive database of 167 studies using a systematic review. In this case the collected data is mostly coded as count data and only a subset of 37 studies was analyzed for actual water yield change. There is some overlap between the two data sets, but there are also some studies unique to both sets. The more regionally concentrated and detailed study by @levy2018 is a further independent dataset with no overlap with the other studies. However, for this study only the flow and rainfall data is available for the catchments, and the change in landcover was derived from satellite data and was not made available.

The conclusions of the first mentioned major review paper [@zhang2017] indicates that there is a distinct difference in the change in flow as a result of forestation or deforestation between small watersheds (catchments), defined as < 1000 km^2^ and large watersheds (catchments) > 1000 km^2^. While for small catchments there was no real change in runoff with changes in cover, for large catchments there was a clear trend showing a decrease in runoff with increases in forest cover. The main conclusion was that the response in annual runoff to forest cover was scale dependent and appeared to be more sensitive to forest cover change in water limited catchments relative to energy limited catchments [@zhang2017].

@@

An earlier paper, that includes much of the same data as @zhang2017 and @filoso2017, is @zhou2015, which has one author in common with @zhang2017. However, this paper aims to explain the variation in the data using the elasticity approach in the Fuh model, which is similar to well-known Budyko approaches [@zhang2018elasticity]. In particular, it aims to link the variation in the observed data to variations in the exponent *m* in the Fuh model. A key observation is that in drier environments, the effects of removing forest cover are much greater than in wetter environments, which is also suggested by Figure 4 in @zhang2017. The Fuh model and the related variations of the Budyko equilibrium modelling approach was also used by @hoekvandijke2022 to interpret the global impact of reforestation.

However, concerning is that there are some clear limitations in these studies, and some of this applies to meta-analyses in general. The main method in the work by @zhang2017 is a single covariate linear regression. In contrast, the systematic review from @filoso2017 mainly emphasises the classification and distributions of the study. @zhang2017 points out that a main assumption in their work is that the catchment size threshold at 1000 km^2^ is a distinct separation between "small" and "large" catchments. However, a subset of 37 data points in @filoso2017 (their Figure 9) does not appear to support this, suggesting a continuum. And while the work @filoso2017 provides important insights in study types, analysis types, forest types and broad classification, there is limited quantification of actual impact.

In contrast to the single covariate linear regression in the earlier studies [@zhang2017; @filoso2017] and the top-down Budyko modelling [@zhou2015; @hoekvandijke2022], the regional Brazilian Cerrado study [@levy2018] provides an example of an carefully designed statistical approach using mixed effects modelling and Differences-in-Differences modelling focusing specifically on the effect of deforestation. The analysis specifically accounted for differences between catchments and differences due to variations in climate. Not all datasets are however suitable for this kind of in-depth analysis.

Given all these previous reviews and the seemingly clear conclusions about the impact of forest cover change on streamflow, the question is why another review paper on this topic?
There is a real attraction in the concept of statistical analysis of past studies encapsulated in meta-analysis to be able to extrapolate findings to larger scales, and to identify factors across global scales [@evaristo2020metaanalysis]. 
However, there are also some hidden complications in this that can invalidate results, which this paper aims to highlight. There are four potential errors (or limitations) in such global meta-analyses:

@@

The aim of this paper is to first reanalyze the global dataset [@zhang2017; @filoso2017] using some more detailed statistical modelling and to use this to highlight examples of each of these limitations. This will show how they have influenced the outcomes of the past work, and provide suggestions of how we can overcome these limitations. In addition, by applying more complex statistical models, we will highlight the conclusions that can be drawn from the data. Finally, we will highlight future research needs in the area of forest cover change impact on streamflow.

@@

As indicated, the starting point of this paper is the data base of studies which were included in @zhang2017 as supplementary material. The columns in this data set are the catchment number, the	catchment name, the	Area in km^2^, the annual average precipitation	(Pa) in mm, the	forest type, hydrological regime, and climate type, the change in forest cover	in \% ($\Delta F\%$) and the change in streamflow in \% $(\Delta Qf\%)$, the precipitation data type, the assessment technique, and the	source of the info, which is a citation. The change in streamflow ($\Delta Qf\%$) is based on equation 1 in @zhang2017.

Several of these columns contain abbreviations to describe the different variables, which are summarised in Table \@ref(tab:table1). These abbreviations will later be used in the models.

@@

The paper by @zhang2017 also uses the dryness index, which is the annual rainfall (Pa) divided by the potential or reference evapotranspiration (ET~0~ or E~0~) in their analysis, and have used the dryness index to identify the climate type. However, the potential or reference ET used for this calculation was originally not included in the published data set. We will discus below how we derived the dryness index in our data set. We combined the tables for small catchments (< 1000 km^2^) and large catchments (>= 1000 km^2^) from @zhang2017 in our analysis.

@@

As highlighted, @zhang2017 did not provide values for evapotranspiration in the data base. Using the location information, reference evapotranspiration (E~0~) was extracted from the Global Aridity Index and Potential Evapo-Transpiration (ET~0~) Climate Databasev2 [@trabucco2018], if a value of E~0~ was not available from the original papers. For large catchments, this value (and the associated coordinates), similar to annual average rainfall, is only an approximation of the climate at the location.

Similar to @zhang2017, the Dryness index was calculated from the catchment estimate of reference evapotranspiration and the catchment estimate of annual average rainfall (Pa) as:

\begin{equation}
Dryness = \frac{E_{0}}{Pa} (\#eq:eq1)
\end{equation}

@@

The aim of the statistical analysis is to highlight the most important variables in the data set that explain the change in flow as a result of changes in forest cover. This first aim is similar to @zhang2017, but the main difference is that we start off with all variables in the data set in the model. Subsequently the analysis will concentrate on how the individual variables in the dataset relate to each other and  how latent variables in the data set can be masked and result in relationships that might not really exist. Finally, the analysis will highlight how the results are conditional on the dataset.

In the statistical analysis we are not necessarily seeking the best "predictive" model, and as such do not perform a traditional variable selection process. Rather, we focus on analyzing the predictor variables in the full model to identify how all the variables explain the variance in the dependent variable.

@@

The changes in forest cover contain both positive (forestation) and negative values (deforestation). In @zhang2017, these changes were jointly analyzed, assuming the effect on the change in flow was linear and the effect of removing forest cover was the same as an equivalent addition of forest cover.

However, the impact of an increase in forest cover can be different from the same fractional decrease in forest cover. The question becomes how best to analyze this. One approach would be to allow a different slope and a different intercept for the decreases relative to the increases. 
This can be tested by converting all the change in forest cover data to positive values, and an additional binary column ($sign_{forest cover}$) can be included indicating whether it was a forest cover increase or decrease. In the model, the parameter for $sign_{forest cover}$ will indicate the difference in the changes in flow for increases in forest cover compared to decreases in forest cover. The disadvantage of this approach is that the relationship with forest cover becomes discontinuous at the origin (0 change in forest cover).

A second approach is to test the change in forest cover as a non-linear relationship in the GAM model. Because a shrinkage penalty is used, this will also test the non-linear assumption and allows the variable for forest cover to be continuous. The disadvantage of this approach is that the relationship between forest cover and change in flow is less easy to interpret, as the non-linear fit in the GAM has no direct parametric form. 
All three approaches are tested in this study.

The overarching test focuses on identifying the change streamflow as a result of a change in forest cover and how this is potentially affected by different other factors (as indicated by the previous research: @zhang2017; @filoso2017; @zhou2015): climate, size of catchment and length of study. In addition to these earlier identified factors, this study also tested for the factors listed in Table \@ref(tab:table1)

@@

The overall dataset contains `r nrow(All_data)` observations of changes in flow, which includes the newly identified data sets and after removing identified duplicate data and lines with missing data. In contrast, the original dataset from @zhang2017 contained `r nrow(Zhang_all)` catchments and the @filoso2017 study used 37 catchments (Table S2 in @filoso2017). The overall distribution of changes in flow is highly skewed as is the distribution of changes in forest cover and *Area km^2^*. The values of changes in flow greater than 100\% and smaller than -100\% clearly create long tails in the change in flow distribution. Note also the large number of studies with 100\% forest cover reduction. Clearly visible is also that smaller catchments dominate the database with 42\% of the data from catchments < 1 km^2^ and 65\% of the data for catchments < 10 km^2^ (Figure 1). This high skew in some of the data can create difficulties in the statistical modelling and this will be discussed later.

```{r, echo = F, warning = F, message=F, eval =F}
png("DataExploration.png", width = 2500, height = 1800, res = 300)
All_data %>%
  mutate(log10Area = log10(Area_km2)) %>%
  pivot_longer(c(DeltaF_perc, DeltaQf_perc, log10Area, Pa_mm, E0), 
               names_to = "variable", values_to = "value") %>%
  ggplot(aes(value)) + geom_histogram(fill = "steelblue", bins=50) + theme_bw() +
  facet_wrap(~variable, scales = "free", labeller = labeller(variable = 
    c("log10Area" =  "log10(Area km2)", #latex2exp::TeX("$\\log_{10}({Area~km^2})$"),
      "DeltaF_perc" = "Delta Forest (%)",
      "DeltaQf_perc" = "Delta Qf (%)",
      "E0" = "Annual Potential E (mm)",
      "Pa_mm" = "Annual Rainfall (mm)"),default = label_parsed)) +   
  theme(axis.title = element_text(size = rel(1.5)),
        axis.text = element_text(size = rel(1.2)),
        strip.text = element_text(size = rel(1.2)),
        axis.text.x = element_text(angle = 45, hjust = 1))
dev.off()
```

@@

```{r correlation, eval=F, echo = F}
All_data2 <- All_data2 %>%
  mutate(Length = To - From,
         mid_year = From + (To - From)/2)

png("variable_corr_plot.png", width = 960, height = 960)
model.matrix(~0+DeltaF_perc + 
                   log10(Area_km2) + 
                    Dryness + 
                    Length +
                    Precip_data_type +  Assessment_technique +
                    Forest_type +
                    Hydrological_regime, data=All_data2) %>% 
  cor(use="pairwise.complete.obs") %>% 
  ggcorrplot(show.diag = F, type="lower", lab=TRUE, lab_size=4)
dev.off()
